---
title: "Map of wall lizard sampling in Switzerland"
output:
  html_document:
    code_folding: hide
    df_print: paged
    theme: cerulean
    toc: true
    toc_float: true
    number_sections: true
    spacing: double
    fontsize: 12pt
    fig_width: 8
    fig_height: 6
date: "2023-06-06"
author: "Anna Orteu, afarreo@gmail.com"
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '.')
```

This is a short tutorial on how to plot maps in R, plotting data onto them such as sampling sites and borders, and how to crop maps using the outline of a country.

For it we will be using sampling data of a project looking at wall lizards in Switzerland, in which diferent locations were sampled in two different years, 2021 and 2023. 

First we set up the working environment, loading required libraries and sampling data, and defining the colour palette.

```{r, message=FALSE, warning=FALSE}
library(ggmap)
library(osmdata)
library(rnaturalearth)
library(sf)
library(raster)
library(maps)

cbPalette <- c("#F0E442","#CC79A7")

#get the sampling data 
data <- read.csv("data/Sample_gps.csv")
data$year = as.factor(data$year)
```

# The sampling sites

We carried out two fieldwork trips, one in 2021 and one in 2023. We sampled multiple locations in 3 Swiss cantons (Aargau, Bern and Ticino). I want to show the sampling sites in a map with particular emphasis on the relief of Switzerland, given that for our study it is relevant that we sampled on both sides of the Alps. 

What does my data look like? 
```{r}
head(data)
```

# Plot the sampling sites

First we plot the sampling sites on a map of the area. We can do that by: 

1. Define the area that we want to plot 
```{r, message=FALSE, warning=FALSE}
#define the area
switzerland <- c(left = 5.7, bottom = 45.6, right = 10.5, top = 47.9)
```

2. Get the background map of the area using `get_map` from the `ggmap` package.
```{r, message=FALSE, warning=FALSE}
#get the map
swiss_map <- get_map(switzerland, source="stamen", maptype = "terrain-background", color = "color", force=TRUE)
```

3. Plot it using the `ggmap` function from the `ggmap` package.
```{r, message=FALSE, warning=FALSE}
#plot it
ggmap(swiss_map)+
  geom_jitter(data=data, aes(x=longitude, y=latitude, fill=year), size=3, shape=21, width = 0.01, height = 0.01)+
  theme(legend.position = "none")+
  scale_fill_manual(values = cbPalette)+
  ylab("Latitude")+
  xlab("Longitude")+
  theme_bw() +
  theme(axis.line = element_line(colour = "grey"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(), legend.position = "top") 
```

# Plot the sampling sites with the outline of Switzerland

Now, we would like to plot on top the outline of Switzerland. 

To do that we first need to get the outline as a sf object using the `ne_countries` function from `rnaturalearth`.
```{r, message=FALSE, warning=FALSE}
#### SWITZERLAND WITH COUNTRY OUTLINE ####
# shape of switzerland from 
swiss_shape <- ne_countries(returnclass = "sf", country = "Switzerland", scale = 'large')
```

Then, we follow the same steps as before to get and plot the map, but this time add the outline of Switzerland using the function `geom_sf` from `ggmap`.
```{r, message=FALSE, warning=FALSE}
#get the map
swiss_map <- get_map(switzerland, source="stamen", maptype = "terrain-background", color = "color", force=TRUE)
#plot it
ggmap(swiss_map)+
  geom_jitter(data=data, aes(x=longitude, y=latitude, fill=year), size=3, shape=21, width = 0.01, height = 0.01)+
  theme(legend.position = "none")+
  scale_fill_manual(values = cbPalette)+
  geom_sf(data=swiss_shape, aes(x = label_x, y = label_y, group = sovereignt), colour = "grey20", 
          fill=NA, linewidth =0.9, linetype="longdash")+
  ylab("Latitude")+
  xlab("Longitude")+
  theme_bw() +
  theme(axis.line = element_line(colour = "grey"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(), legend.position = "top") 
```



# Plot the sampling sites with the outline of Switzerland and canton borders

We can do the same but also add the canton borders in the plot. Looks very cluttered but could be useful in some cases.

```{r, message=FALSE, warning=FALSE}
#### SWITZERLAND WITH CANTONS ####
#get tge borders
swiss_shape <- ne_states(country = "Switzerland", returnclass = "sf")
#get the map
swiss_map <- get_map(switzerland, source="stamen", maptype = "terrain-background", color = "color", force=TRUE)
#plot it
ggmap(swiss_map)+
  geom_jitter(data=data, aes(x=longitude, y=latitude, fill=year), size=3, shape=21, width = 0.01, height = 0.01)+
  theme(legend.position = "none")+
  scale_fill_manual(values = cbPalette)+
  geom_sf(data=swiss_shape, aes(x = longitude, y = latitude, group = adm0_a3), colour = "grey30", 
          fill=NA, linewidth =0.9, linetype="longdash")

```

# Plot the sampling sites in a cutout of the country

Now, we would like to generate a similar plot of the sampling sites, but cutting out any part of the map outside Switzerland. That is we want to generate a cut-out of Switzerland where we can plot our data. 

To do this we first need to transform the base map to a raster using a custom function courtesy R Lovelace from https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/download-raster.R. 

Then, we can cutout the silhouette of Switzerland using the `crop` and `mask` funstions from the `raster` package. 

```{r, message=FALSE, warning=FALSE}
#### CROP MAP WITH OUTLINE #####

# courtesy R Lovelace from https://github.com/Robinlovelace/Creating-maps-in-R/blob/master/vignettes/download-raster.R
ggmap_rast <- function(map){
  map_bbox <- attr(map, 'bb') 
  .extent <- extent(as.numeric(map_bbox[c(2,4,1,3)]))
  my_map <- raster(.extent, nrow= nrow(map), ncol = ncol(map))
  rgb_cols <- setNames(as.data.frame(t(col2rgb(map))), c('red','green','blue'))
  red <- my_map
  values(red) <- rgb_cols[['red']]
  green <- my_map
  values(green) <- rgb_cols[['green']]
  blue <- my_map
  values(blue) <- rgb_cols[['blue']]
  stack(red,green,blue)
}

che.rast <- ggmap_rast(map = swiss_map)

che.rast2 <- crop(che.rast, swiss_shape) 
che.rast3 <- mask(che.rast2, swiss_shape)
```

Once we have our cropped raster, we transform it to points in a dataframe to be able to plot it with `ggplot2` and create the base map.

```{r, message=FALSE, warning=FALSE}
# prep raster as a data frame for printing with ggplot
che.df <- data.frame(rasterToPoints(che.rast3))
che_map <- ggplot(che.df) + 
  geom_point(aes(x=x, y=y, col=rgb(layer.1/255, layer.2/255, layer.3/255))) + 
  scale_color_identity()
```

Then we plot the map using `coord_map()` to make sure that out plot maintains the mercator projection and add the data points to it.

```{r}
# With correct mercator projection
che_map + 
  coord_map() +
  geom_jitter(data=data, aes(x=longitude, y=latitude, fill=year), size=3, shape=21, width = 0.01, height = 0.01)+
  theme(legend.position = "none")+
  scale_fill_manual(values = cbPalette)+
  ylab("Latitude")+
  xlab("Longitude")+
  theme_void() +
  theme(legend.position = c(0.9, 0.85), 
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        plot.title = element_text(size=20, face="bold", hjust = 0.5)) +
  guides(fill=guide_legend(title="Sampling year"))+
  ggtitle("Wall lizard sampling in Switzerland")


```


